---
title: "Assignment 3"
format:
  html:
    code-fold: false
    embed-resources: true
    toc: true
---

# Dataset Description: U.S. Bilateral Trade in Poultry Products  

This dataset contains bilateral international trade data for the United States as the reporter country, focusing on the commodity classified under "**meat and edible offal**" of poultry, traded in fresh, chilled, or frozen form.  

The dataset includes poultry products, covering common cuts and parts such as breasts, thighs, wings, livers, hearts, and gizzards. It excludes live poultry and processed poultry products.  

Trade flows are reported with all countries as partners, meaning the data capture U.S. poultry trade with every trading partner (countries) worldwide. The dataset includes:  

-   Imports: Poultry products entering the United States from foreign suppliers (e.g., frozen chicken from Brazil or Canada).  

-   Exports: Poultry products shipped from the United States to international markets (e.g., U.S. poultry exports to Mexico or Asian countries).  

The data are aggregated at an annual frequency and cover the period 2018–2025.  

# Data Source:  

The dataset is obtained from the UN Comtrade database, an open-access repository of official international trade statistics compiled from customs data reported by over 150 countries. Key variables include trade values (USD) and quantities (kilograms), reflecting officially reported bilateral trade flows between the United States and its trading partners.  

# General directions  

1.  For each question, you are encouraged to create a chunk below it, and fully develop the code that creates your answer.  

2.  You are encouraged to use wrangling functions from the packages `dplyr` and `tidyr`, and plotting functions from the package `ggplot2`.  

3.  You are encouraged to the the pipe operator (`%>%` or `|>`) to pipe wrangling steps together.  

4.  For each question, after showing your code and output (be it numerical or graphical), you also need to provide a written sentence of your answer. Just showing code output is not sufficient for full points.  

5.  As part of the initial wrangling, you are expected to:  

-   Transform the `qty` column from kg to Mg  
-   Select only columns that are useful to answer the questions.  

6. Check out the following cheatsheets to give you some ideas on wrangling and plotting:  

  - **ggplot2**: https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf  
  - **dplyr**: https://rstudio.github.io/cheatsheets/data-transformation.pdf   
  


## Dataset

```{r}

library(readxl) # to read excel files
library(janitor) # to clean data
library(dplyr) # wrangling
library(tidyr) # wrangling
library(readr) # to export csv
library(ggplot2)
```

```{r}

chicken <- read_csv("03_assignment_data.csv")

# Checking first rows
chicken


glimpse(chicken)

```

```{r Summary}
summary(chicken)
```


# Assignemnt Questions:

## Imports:

#### 1. How many distinct countries were there (based on given data), from where US was importing the poultry products from 2018-2024?

```{r Clean}
chicken_import <- clean_names(chicken)
chicken_import <- filter(chicken_import, flow_desc == "Import")
chicken_import

```
```{r}
#count unique values in partner column
unique_countries <- n_distinct(chicken_import$partner_desc)
unique_countries

```
##### ***There were 7 unique countries the U.S. was importing from during the 2018-2024 period.***

#### 2. What are the minimum, maximum trade values for imports? Mention the country, year and its respective value.
```{r}
#mutate import qty from kg to Mg
import_01 <- mutate(chicken_import,
                  qty_mg = qty*1000000,
                  qty_mg= round(qty_mg, 0)
                  )
import_01
```


```{r}
#summarize for mean, minimum, maximum, etc. 
import_sum <- select(import_01, ref_year, period, flow_desc,
 partner_desc, qty_unit_abbr, qty, primary_value, qty_mg
)  
import_sum


import_sort <- arrange(import_sum, primary_value)
import_sort

```
#####***The mininum value for trade imports came from Poland in 2022 at 4308 (primary value based on FOB or CIF). The maximum value came from Chile in 2022 at 428214676 primary value.***
#### 3. Identify the top 5 countries from which the USA imported the most (qty) in the years 2018 and 2024. List the countries and their corresponding import values.
```{r}

```

#### 4. Calculate the total import value for the USA from each country. Which country had the highest total export value **to** US, and what was that value?
```{r}

```

#### 5. Calculate the total import value for the USA in each year from 2018 to 2025. Which year had the highest total import value, and what was that value?
```{r}

```

#### 6. Plot a line graph showing the total annual import value for the USA from 2018 to 2024. Describe any noticeable trends or patterns (e.g., increases, decreases, or fluctuations) in the graph. Create a publication ready plot.
```{r}

```

## Exports:

#### 7. How many distinct countries were there (based on given data), to where US was exporting the poultry products?
```{r}

```

#### 8. What are the minimum, maximum trade values for exports? Mention the country, year and its respective value.
```{r}

```

#### 9. Identify the top 5 countries to which the USA exported the most (qty) in the years 2018 and 2024. List the countries and their corresponding export values.
```{r}

```

#### 10. Calculate the total export value for the USA from each country. Which country had the highest total import value **from** US, and what was that value?
```{r}

```

#### 11. Calculate the total export value for the USA in each year from 2018 to 2025. Which year had the highest total export value, and what was that value?
```{r}

```

#### 12. Plot a line graph showing the total annual export value for the USA from 2018 to 2024. Describe any noticeable trends or patterns (e.g., increases, decreases, or fluctuations) in the graph. Create a publication ready plot.
```{r}

```

## Imports - Exports Comparision:

#### 13. Does the U.S. behave more as a net importer or net exporter for this commodity? Explore it:

-   Based on total value US imports and exports.
-   Based on number of countries US imports and exports
-   Based on each year's import export value.
```{r}

```

#### 14. Are the main import partners also major export destinations? Explore it based on the total import export value by countries.

#### 15. Can you calculate if US is in trade surplus or trade deficit for each year from 2018-2024?

$$\text{Trade surplus or deficit} = \text{Imports} - \text{Exports}$$
```{r}

```

# Submitting your work  

Once you have developed all the code and answers, make sure to Render this quarto file.  

**Notes on rendering**:  

- Make sure to render your work and inspect how the final html look like.  
- If it does not look professional for whatever reason, then fix the issue, re-render it, recheck.  
- Only send me your work once your html file looks professional.  
- Some potential issues you may encounter and how to fix them:   
  - Some times your code may be creating a table output that is waaay to long and cumbersome to scroll through when rendered. If this is the case, make it more professional looking by using the `head()` function to only print the first handful of rows (instead of thousands of rows).  
  
  - **DO NOT** delete the file's heading levels (# and ##). They set up the proper heading 1 and 2 levels, and I use them to guide my grading.  
  
  - There's a total of **15 questions**. If you mess around with the heading levels and you get a different number of questions on the rendered version, fix it.  
  
  - If a given chunk is also outputting warnings or messages, inhibit this behavior by changing the chunk options `message` and `warning` to `FALSE` (I'll demo this in a moment).  
  
  - If, after rendered, 2 lines of text are connected and you wish to "break line" between them, add 2 extra spaces after the first one (I'lld demo this in a moment).    
  
After rendering, an .html file will be created on your `code` folder.  

Rename this file to `03_assignment_LASTNAME.html`.    

For ex., `03_assignment_Bastos.html`.

Submit the html file on eLC by Monday Feb 2nd 11:59 pm, **one week** from today.  

